# `wrassp` {#app-chap:wrassp}

## Using Praat's signal processing routines in the EMU-SDMS {#sec:app-chap-wrassp-praatsSigProc}

### `To Formant (burg)...` to SSFF files

The R code snippet below shows how generating an `AsspDataObj` from scratch can be used in a function to place data from other sources into SSFF files. In this case it uses the `PraatR` R package (see http://www.aaronalbin.com/praatr/index.html) to execute Praat's `"To Formant (burg)..."` function to then store the data to a comma separated file using `"Down to Table..."`. The generated table is then read into R and the appropriate columns are placed into tracks of a `AsspDataObj` object. The `PraatToFormants2AsspDataObj` can be viewed as a template function as it can easily be adapted to use other functions provided by Praat or even other external tools.

NOTE: this function can be accessed directly: `source("https://raw.githubusercontent.com/IPS-LMU/The-EMU-SDMS-Manual/master/R/praatToFormants2AsspDataObj.R")`


```{r, eval=F, code=readLines("R/praatToFormants2AsspDataObj.R")}
```

How this function can be applied to wav files of an emuDB is shown below.

```{r, eval=F, message=F}
library(emuR)

# create demo data in tempdir()
create_emuRdemoData(tempdir())

# create path to demo database
path2ae = file.path(tempdir(), "emuR_demoData", "ae_emuDB")

# list all .wav files in the ae emuDB
paths2wavFiles = list.files(path2ae, pattern = "*.wav$", 
                            recursive = TRUE, full.names = TRUE)

# loop through files
for(fp in paths2wavFiles){
  ado = praatToFormants2AsspDataObj(fp)
  newPath = paste0(tools::file_path_sans_ext(fp), '.praatFms')
  # print(paste0(fp, ' -> ', newPath)) # uncomment for simple log
  wrassp::write.AsspDataObj(ado, file = newPath)
}

# load emuDB
# (verbose = F is only set to avoid additional output in manual)
ae = load_emuDB(path2ae, verbose = FALSE)

# add SSFF track definition
add_ssffTrackDefinition(ae, 
                        name = "praatFms", 
                        columnName = "fm",
                        fileExtension = "praatFms")

# test query + get_trackdata
sl = query(ae, "Phonetic == n")

# (verbose = F is only set to avoid additional output in manual)
td = get_trackdata(ae, 
                   sl, 
                   ssffTrackName = "praatFms", 
                   verbose = F)
```

```{r echo=FALSE, results='hide', message=FALSE}
# clean up emuR_demoData
unlink(file.path(tempdir(), "emuR_demoData"), recursive = TRUE)
```

### `To Pitch...` to SSFF files

The R code snippet below does the following:

a) calculates f0 via Praat's "To Pitch..." command;
b) smooths the f0 contour via Praat's "Smooth" command, if parameter "smooth" is set to TRUE
c) creates a PitchTier and then a TableOfReal
d) converts all of this into an AsspDataObj (which later on can be saved as an SSFF-file.)

A few comments about synchronized F0 and Formant values, whenever `praatToFormants2AsspDataObj()` and
`praatToPitch2AsspDataObj()` are involved:

- Be careful with the arguments list: 
- **The first entry is "Time step (s) (standard value: 0.0) the measurement interval (frame duration), in seconds.**
If you supply 0, Praat will use a time step of 0.75 / (pitch floor), e.g. 0.01 seconds if the pitch floor is 75 Hz;
in this example, Praat computes 100 pitch values per second. "
Here parameter 1 - Time step - is set to 0.00625 (Seconds) (as opposed to Praat's default of 0.0) in order
to keep it in line with Time step in "To Formant...",
because Time step in "To Formants..." is derived from
window length (which in "To Formant..." defaults to 0.025) (window length/4 (--> Time step in "To Formant..."
will usually be 0.00625))
- **The second parameter "Pitch floor (Hz) (standard value: 75 Hz)**:
candidates below this frequency will not be recruited.
This parameter determines the length of the analysis window: it will be 3 longest periods long, i.e.,
if the pitch floor is 75 Hz, the window will be 3/75 = 0.04 seconds long.
Note that if you set the time step to zero, the analysis windows for consecutive measurements will overlap appreciably:
Praat will always compute 4 pitch values within one window length, i.e., the degree of oversampling is 4."
Importantly, this parameter is set NOT to praat's default 75 Hz, but to 60 Hz, again because of correspondance
of window lengths between "To Pitch..." and "To Formants...". The actual window length in "To Formants..." will be twice as long
as the value given in the "To Formants..." command, i.e. the default of 0.025 will result in a window length of 0.05.
A window length in "To Pitch..." can indirectly achieved by using a pitch floor value of 60 Hz (given that 3/60 = 0.05).
In most cases, differing window lengths will not affect the temporal position of the F0 and Formant values, however, due
to problems near the edges, sometimes they will (and therefore result in non-synchronized F0 and Formant values).
Due to rounding errors, F0 and Formant values still might be slightly asynchronous; to avoid this, `praatToPitch2AsspDataObj()`
rounds the start time with a precicion of 0.001 ms (via round(attr(ado, "startTime"),6) in the very end).
- **The third parameter (default: 600) is the pitch ceiling** (and this parameter will not affect any other parameters indirectly)

NOTE: this function can be accessed directly: `source("https://raw.githubusercontent.com/IPS-LMU/The-EMU-SDMS-Manual/master/R/praatToPitch2AsspDataObj.R")`


```{r, eval=F, code=readLines("R/praatToPitch2AsspDataObj.R")}
```

How this function can be applied to wav files of an emuDB is shown below.

```{r, eval=F, message=F}
library(emuR)

# create demo data in tempdir()
create_emuRdemoData(tempdir())

# create path to demo database
path2ae = file.path(tempdir(), "emuR_demoData", "ae_emuDB")

# test the function of converting praat formant data to emuR
paths2wavFiles = list.files(path2ae,
                            pattern = "wav$",
                            recursive = TRUE,
                            full.names = TRUE)

# loop through files
for(fp in paths2wavFiles){
  ado = praatToPitch2AsspDataObj(fp)
  newPath = paste0(tools::file_path_sans_ext(fp), '.praatF0')
  # print(paste0(fp, ' -> ', newPath)) # uncomment for simple log
  wrassp::write.AsspDataObj(ado, file = newPath)
}

# load emuDB
# (verbose = F is only set to avoid additional output in manual)
ae = load_emuDB(path2ae, verbose = FALSE)

# add SSFF track definition
add_ssffTrackDefinition(ae,
                        name = "praatF0",
                        columnName = "f0",
                        fileExtension = "praatF0")

# test query + get_trackdata
sl = query(ae, "Phonetic == n")
# (verbose = F is only set to avoid additional output in manual)
td = get_trackdata(ae, 
                   sl, 
                   ssffTrackName = "praatF0", 
                   verbose = F)

# configure EMU-webApp to show new track
sc_order = get_signalCanvasesOrder(ae, "default")

set_signalCanvasesOrder(ae, "default", c(sc_order, "praatF0"))

# serve(ae) # uncomment to view in EMU-webApp
```

```{r echo=FALSE, results='hide', message=FALSE}
# clean up emuR_demoData
unlink(file.path(tempdir(), "emuR_demoData"), recursive = TRUE)
```



## Using OpenSMILE signal processing routines in the EMU-SDMS {#sec:app-chap-wrassp-opensmileSigProc}

NOTE: this function can be accessed directly as follows: `source("https://raw.githubusercontent.com/IPS-LMU/The-EMU-SDMS-Manual/master/R/SMILExtract2AsspDataObj.R")`

```{r, eval=FALSE, code=readLines("R/SMILExtract2AsspDataObj.R")}
```

How this function can be applied to wav files of an emuDB is shown below.

```{r, eval=F, message=F}
library(emuR)

# create demo data in tempdir()
create_emuRdemoData(tempdir())

# create path to demo database
path2ae = file.path(tempdir(), "emuR_demoData", "ae_emuDB")

# list all .wav files in the ae emuDB
paths2wavFiles = list.files(path2ae,
                            pattern = "*.wav$",
                            recursive = TRUE,
                            full.names = TRUE)

# loop through files
for(fp in paths2wavFiles){
  ado = SMILExtract2AsspDataObj(fp,
                                SMILExtractPath = "~/programs/opensmile-2.3.0/bin/SMILExtract",
                                configPath = "~/programs/opensmile-2.3.0/config/demo/demo1_energy.conf")
  newPath = paste0(file_path_sans_ext(fp), '.SMILExtract')
  # print(paste0(fp, ' -> ', newPath)) # uncomment for simple log
  write.AsspDataObj(ado, file = newPath)
}

# load emuDB
# (verbose = F is only set to avoid additional output in manual)
ae = load_emuDB(path2ae, verbose = FALSE)

# add SSFF track definition
add_ssffTrackDefinition(ae,
                        name = "SMILExtract",
                        columnName = "pcm_LOGenergy",
                        fileExtension = "SMILExtract")

# test query + get_trackdata
sl = query(ae, "Phonetic == n")
# (verbose = F is only set to avoid additional output in manual)
td = get_trackdata(ae,
                   sl,
                   ssffTrackName = "SMILExtract",
                   verbose = F)

# test display
set_signalCanvasesOrder(ae,
                        perspectiveName = "default",
                        order = c("OSCI", "SPEC", "SMILExtract"))

# serve(ae) # uncomment to view in EMU-webApp
```

```{r echo=FALSE, results='hide', message=FALSE}
# clean up emuR_demoData
unlink(file.path(tempdir(), "emuR_demoData"), recursive = TRUE)
```

